---
title: "NHL Covid Impact ECON 1700"
author: "Ian Keller"
date: "2024-04-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Reading in data

```{r}
#2016 Season
nhl_2016 <- read.csv("nhl_2016.csv")
nhl_2016_playoffs <- read.csv("nhl_2016_playoffs.csv")

#2017 Season
nhl_2017 <- read.csv("nhl_2017.csv")
nhl_2017_playoffs <- read.csv("nhl_2017_playoffs.csv")

#2018 Season
nhl_2018 <- read.csv("nhl_2018.csv")
nhl_2018_playoffs <- read.csv("nhl_2018_playoffs.csv")

#2019 Season
nhl_2019 <- read.csv("nhl_2019.csv")
nhl_2019_playoffs <- read.csv("nhl_2019_playoffs.csv")

#2020 Season
nhl_2020 <- read.csv("nhl_2020.csv")
nhl_2020_playoffs <- read.csv("nhl_2020_playoffs.csv")

#2021 Season
nhl_2021 <- read.csv("nhl_2021.csv")
nhl_2021_playoffs <- read.csv("nhl_2021_playoffs.csv")

#May not include the following data after covid.

#2022 Season
nhl_2022 <- read.csv("nhl_2022.csv")
nhl_2022_playoffs <- read.csv("nhl_2022_playoffs.csv")

#2023 Season
nhl_2023 <- read.csv("nhl_2023.csv")
nhl_2023_playoffs <- read.csv("nhl_2023_playoffs.csv")
```

### Creating playoff variable

```{r}
# Playoff = 1
nhl_2016_playoffs$playoffs = 1
nhl_2017_playoffs$playoffs = 1
nhl_2018_playoffs$playoffs = 1
nhl_2019_playoffs$playoffs = 1
nhl_2020_playoffs$playoffs = 1
nhl_2021_playoffs$playoffs = 1
nhl_2022_playoffs$playoffs = 1
nhl_2023_playoffs$playoffs = 1

#Playoff = 0
nhl_2016$playoffs = 0
nhl_2017$playoffs = 0
nhl_2018$playoffs = 0
nhl_2019$playoffs = 0
nhl_2020$playoffs = 0
nhl_2021$playoffs = 0
nhl_2022$playoffs = 0
nhl_2023$playoffs = 0
```
###Creating Season Variable
```{r}
nhl_2016_playoffs$season = 2016
nhl_2017_playoffs$season = 2017
nhl_2018_playoffs$season = 2018
nhl_2019_playoffs$season = 2019
nhl_2020_playoffs$season = 2020
nhl_2021_playoffs$season = 2021
nhl_2022_playoffs$season = 2022
nhl_2023_playoffs$season = 2023

#Playoff = 0
nhl_2016$season = 2016
nhl_2017$season = 2017
nhl_2018$season = 2018
nhl_2019$season = 2019
nhl_2020$season = 2020
nhl_2021$season = 2021
nhl_2022$season = 2022
nhl_2023$season = 2023
```



### Merging Data

```{r}
# Merge each season with playoffs
merged_2016 <- rbind(nhl_2016, nhl_2016_playoffs)
merged_2017 <- rbind(nhl_2017, nhl_2017_playoffs)
merged_2018 <- rbind(nhl_2018, nhl_2018_playoffs)
merged_2019 <- rbind(nhl_2019, nhl_2019_playoffs)
merged_2020 <- rbind(nhl_2020, nhl_2020_playoffs)
merged_2021 <- rbind(nhl_2021, nhl_2021_playoffs)
merged_2022 <- rbind(nhl_2022, nhl_2022_playoffs)
merged_2023 <- rbind(nhl_2023, nhl_2023_playoffs)

#merging all seasons
merged_all <- rbind(merged_2016, merged_2017, merged_2018, merged_2019, merged_2020, merged_2021, merged_2022, merged_2023)
#Merging all the data up to Covid-19
merged_covid <- rbind(merged_2016, merged_2017, merged_2018, merged_2019, merged_2020, merged_2021)

#merging all of the playoff games
merged_playoffs <- rbind(nhl_2016_playoffs, nhl_2017_playoffs, nhl_2018_playoffs, nhl_2019_playoffs, nhl_2020_playoffs, nhl_2020_playoffs)
```

### Cleaning Data

Converting NA's to Zeroes

```{r}
merged_covid[is.na(merged_covid)] <- 0
```

###Creating Distance Data for Team Distances

```{r}
nhl_distance <- read.csv("nhl_distance.csv", header = TRUE, sep = ",")
nhl_distance <- nhl_distance[1:33, 1:33]
#Cleaning data
nhl_distance[nhl_distance == "-"] <- 0
nhl_distance <- nhl_distance[-nrow(nhl_distance), ]

#Changing Row and Column names
new_column_names <- nhl_distance[, 1]
new_column_names[14] = "Los Angeles Kings"

rownames(nhl_distance) <- new_column_names
nhl_distance <- nhl_distance[, -1]
colnames(nhl_distance) <- new_column_names

```

Creating Binary values

```{r}
nhl_distance_numeric <- as.data.frame(lapply(nhl_distance, as.numeric))
distance_final <- ifelse(nhl_distance_numeric < 300, 1, 0)
colnames(distance_final) <- new_column_names
rownames(distance_final) <- new_column_names
```

Removing Individual Team from matrix diagonal

```{r}
# Get the number of rows and columns
n <- nrow(distance_final)
m <- ncol(distance_final)

# Loop through each element in the diagonal and set it to 0
for (i in 1:min(n, m)) {
  distance_final[i, i] <- 0
}
```

Combined Names

```{r}
#rownames(distance_final)[1]
#distance_final[1,]
distance_final <- as.data.frame(distance_final)

# Initialize an empty vector to store the messages
close_games <- c()

# Iterate over each row
for (i in 1:nrow(distance_final)) {
  # Check if the row contains a 1
  if (1 %in% distance_final[i, ]) {
    # Find the column name where 1 exists
    col_name <- names(distance_final)[which(distance_final[i, ] == 1)]
    # Construct the message
    message <- paste(rownames(distance_final)[i], "@", col_name)
    # Append the message to the vector
    close_games <- c(close_games, message)
  }
}

# Print all captured lines
print(close_games)



```

Creating Team versus Team String Variables

```{r}
merged_covid$matchup <- paste(merged_covid$Visitor, merged_covid$Home, sep = " @ ") 
```

Iterating through matchups to see if it is contained within close games

```{r}
# Initialize an empty list to store 1's and 0's
close_list <- c()

# Iterate over each row
for (i in 1:nrow(merged_covid)) {
  if (merged_covid$matchup[i] %in% close_games) {
    close_list <- c(close_list, 1)
  } else {
    close_list <- c(close_list, 0)
  }
}

merged_covid <- cbind(merged_covid, close_matchup = close_list)
```

```{r}
paste("Number of Close games:", sum(merged_covid$close_matchup))
```

###Variable Creation ####Creating Variables for home differential and a binary variable for if there were fans or not(0 if fans yes, 1 if fans no). Here, I will just be looking at the 5 regular seasons before Covid along with Covid impacted seasons.

```{r}
#All Games
merged_covid$home_difference <- merged_covid$G.1 - merged_covid$G
merged_covid$no_crowd <- ifelse(merged_covid$Att. == 0, 1, 0)
merged_covid$home_wins <- ifelse(merged_covid$G.1 > merged_covid$G, 1, 0)
```

###Running the linear regression model This first model looks at a holistic view of the NHL seasons from 2015-16 to 2020-2021.

```{r}
model <- lm(home_difference ~ no_crowd, data = merged_covid)

summary(model)
anova(model)


model2 <- lm(home_difference ~ no_crowd, data = merged_covid[merged_covid$season %in% c(2020,2021), ])
summary(model2)

```

Next, I will look at the impact on playoffs as crowds tend to be more full and excited for rivalry and important games. It is important to note the bias in the playoffs that in a 7 game series the "better team" based on regular season standings has more home games then the away team.

### Simple Playoff Model
```{r}
#cleaning data and creating variables
merged_playoffs[is.na(merged_playoffs)] <- 0
#Playoff Games
merged_playoffs$home_difference <- merged_playoffs$G.1 - merged_playoffs$G
merged_playoffs$no_crowd <- ifelse(merged_playoffs$Att. == 0, 1, 0)


model_playoffs <- lm(home_difference ~ no_crowd, data = merged_playoffs)
summary(model_playoffs)
```

### Looking at the 2021 regular season as teams only played in conference. Fixed effects for matchup?

```{r}
#Cleaning data and adding variables
nhl_2021[is.na(nhl_2021)] <- 0
nhl_2021$home_difference <- nhl_2021$G.1 - nhl_2021$G
nhl_2021$no_crowd <- ifelse(nhl_2021$Att. == 0, 1, 0)


model_2021 <- lm(home_difference ~ no_crowd, data = nhl_2021)
summary(model_2021)
```

### Creating Playoff and Crowd interaction variable

```{r}
model_interaction <- lm(home_difference ~ no_crowd + playoffs + no_crowd*playoffs, data = merged_covid)

summary(model_interaction)
anova(model_interaction)
```

### Full Model

```{r}
full_model <- lm(home_difference ~ no_crowd + close_matchup + playoffs + no_crowd*playoffs + no_crowd*close_matchup, data = merged_covid)
summary(full_model)
```
### Full Model 2020-2021
```{r}
full_model_2021 <- lm(home_difference ~ no_crowd + close_matchup + playoffs + no_crowd*playoffs + no_crowd*close_matchup, data = merged_covid[merged_covid$season %in% c(2020,2021), ])
summary(full_model_2021)
```



### Fixed Effects
```{r}
library(lfe)
fe_reg <- felm(home_difference ~ no_crowd + no_crowd*playoffs | 
                 matchup, data = merged_covid) #[merged_covid$season %in% c(2020,2021), ])

summary(fe_reg)
```
Ran into multicollinearity when modeling with close_game and fe of matchup, since within certain matchups it always will be a close game or not.

```{r}
library(lfe)
fe_reg_covid <- felm(home_difference ~ no_crowd + no_crowd*playoffs | 
                 matchup, data = merged_covid[merged_covid$season %in% c(2020,2021), ])

summary(fe_reg_covid)
```
Just looking at crowd fixed on matchup
```{r}
library(lfe)
fe_reg_crowd <- felm(home_difference ~ no_crowd | 
                 matchup, data = merged_covid[merged_covid$season %in% c(2020,2021), ])

summary(fe_reg_crowd)
```





### Just looking at close matchups

```{r}
model_close <- lm(home_difference ~ no_crowd + close_matchup + no_crowd*close_matchup, data = merged_covid)
summary(model_close)
```

### Exploring Logistic Regression

```{r}
log_model <- glm(home_wins ~ no_crowd + close_matchup + playoffs + no_crowd*playoffs + no_crowd*close_matchup, family = "binomial", data = merged_covid)
summary(log_model)
```
Interpreting Full Model
```{r}
full_coef <- coef(log_model)
interp <- NULL
for (i in 1:length(full_coef)) {
  coef_i <- exp((full_coef)[[i]]) - 1
  coef_good <- coef_i *100
  coef_round <- round(coef_good,4)
  results <- paste("beta", i-1, "interprets as a", coef_round, "% increase/decrease")
  interp <- c(interp, results)
}

interp
```





Simple log model

```{r}
simple_log <- glm(home_wins ~ no_crowd, data = merged_covid, family = "binomial")
summary(simple_log)
```
Interpreting Logistic Coefficients
```{r}
intercept = exp(coef(simple_log)[[1]]) - 1
print(intercept)

coef_b1 <- exp(coef(simple_log)[[2]]) - 1
print(coef_b1)

```




Summing Playoffs and No crowd

```{r}
# Filter rows where playoffs and no_crowd equal 1
subset_df <- subset(merged_covid, close_matchup == 1 & no_crowd ==1)

nrow(subset_df)
```

### Summary Statistics
```{r}
summary(merged_covid)
```

```{r}
library(ggplot2)

histogram <- ggplot(merged_covid, aes(x = home_difference)) +
  geom_histogram(binwidth = 1, fill = "lightgrey", color = "black") +  # Adjust binwidth as needed
  labs(title = "Distribution of Home Difference",
       x = "Home Difference",
       y = "Frequency") +
  theme_minimal()

ggsave("histogram.png", plot = histogram, width = 6, height = 4, dpi = 300)
```


```{r}
library(ggplot2)

# Assuming merged_covid is your dataset and no_crowd is the binary variable of interest

# Calculate average home_difference for each level of no_crowd
avg_home_difference <- aggregate(home_difference ~ no_crowd, data = merged_covid, FUN = mean)

# Create a bar chart with black border
bar_chart <- ggplot(avg_home_difference, aes(x = factor(no_crowd), y = home_difference)) +
  geom_bar(stat = "identity", fill = "lightgrey", color = "black") +  # Adjust fill color and border color as needed
  labs(title = "Average Home Difference by Crowd Presence",
       x = "0 if Crowd, 1 if No Crowd",
       y = "Average Home Difference") +
  theme_minimal()

# Display the bar chart
bar_chart

```
Avg number of fans

```{r}
# Assuming 'merged_covid' is your data frame
# Assuming 'season' and 'Att.' are the column names in your data frame

# Create a histogram for each season
par(mfrow = c(2, 3)) # Adjust layout to fit multiple plots in one window

# Iterate over unique seasons
for (s in unique(merged_covid$season)) {
  # Subset data for the current season
  season_data <- merged_covid$Att.[merged_covid$season == s]
  
  # Plot histogram with adjusted scale and breaks
  hist(season_data, main = paste("Histogram of Att. for Season", s),
       xlab = "Att.", ylab = "Frequency", breaks = 10, xlim = c(0, 25000)) # Adjust breaks and xlim as needed
}

# Reset layout to default
par(mfrow = c(1, 1))

```


```{r}
hist(merged_covid$Att.)
```

Simple df
```{r}
library(dplyr)

merged_simple_final <- merged_covid %>%
  select(-X, -Notes, -LOG)
```


```{r}
summary_table <- summary(merged_simple_final)

# Assuming 'merged_covid' is your data frame containing your variables
print(summary_table)

```












